---
title: "Introduction To R"
output: pdf_document
---

# **Establishing our Basics**

***What is programming?***

Imagine you are a tutor helping a friend learn a new language. Your friend, though very smart, doesn't understand English, so you need to use a special language they do understand. This language is like a programming language, and your friend is the computer.

***What is R?***

R is a programming language designed for statistical computing and data analysis.

**Terms that might show up**

Compiler: converts the stuff you write to stuff machines can understand

# **Basic R Syntax**

1.  [***Comments***]{.underline}: These are used to explain what is going on and ignored by the R interpreter. Comments are preceded the '\#' symbol

    ```{r}
    # This is a comment
    ```

2.  [***Assignment***]{.underline}: You can use the '`<-`' or '`=`' symbol for setting variables equal to something.

    ```{r}
    x <- 5 
    x = 5
    ```

3.  [***Print Output:***]{.underline} The '`print()`' function is used to display the value of a variable.

    ```{r}
    print(x)
    ```

4.  [***Data Types:***]{.underline}

    1.  *Numeric*: represent real numbers (integers and decimals)

    2.  *Character*: represent characters or strings (sequence of characters)

    ```{r}
    #character variable, you can use either ' ' or " " to surround the characters 
    char_var = 'Hello World'
    char_var2 = "Hello World" 

    print(char_var)
    print(char_var2)

    ```

    3.  *Logical*: represents boolean values (i.e. 'TRUE' or 'FALSE')

    ```{r}
    boolean_var = 5 > 6

    print(boolean_var)
    ```

    4.  *Vector*: one-dimensional array that can hold element of the SAME data type

    ```{r}
      # Numeric vector
      numeric_vector <- c(1, 2, 3, 4, 5)
      print(numeric_vector)
      
      # Character vector
      char_vector <- c("apple", "grapes", "banana")
      
      print(char_vector)
    ```

    5.  [***Other Data Types (You can skip this :) )***]{.underline}

    -   *Matrix*: a matrix is a 2D array with rows and columns

    ```{r}
    # Matrix
    matrix_var <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3)

    print(matrix_var)
    ```

    -   *Array*: multi-dimensional generalization of a matrix

    ```{r}
    # Creating a 3D array 
    array_var <- array(c(1, 2, 3, 4, 5, 6), dim = c(2, 2, 2))

    # Displaying the array
    print(array_var)

    ```

    -   *List*: collection of different data types (e.g. characters, vectors, numeric, etc.)

    ```{r}
    # List
    list_var <- list(name = "John", age = 25, is_student = TRUE)

    print(list_var)
    ```

    -   Data Frame: A data fram is a 2D dimensional table with rows and columns where each column can be a different data type

    ```{r}
    # Data frame
    df_var <- data.frame(name = c("Alice", "Bob", "Charlie"), age = c(25, 30, 22))

    print(df_var)
    ```

    6.  [***Indexing***]{.underline}: R uses 1-base indexing meaning the index of the first element in a vector is 1.

```{r}
numeric_vector[1]  # Access the first element
char_vector[2]    # Access the second element

```

7.  [***Functions***]{.underline}: functions are defined using the 'function()' keyword. The 'return()' statement is used to return a value from a function

```{r}
# Example of a function that adds two numbers together
add_numbers <- function(a, b) {
  result <- a + b
  return(result)
}

# Call the function
sum_result <- add_numbers(3, 5)
print(sum_result)

```

# **Using R for Statistics**

## **Descriptive Statistics**

| Statistics Concept        | Function in R | parameters                            |
|-----------------------|------------------|-------------------------------|
| Mean                      | mean(x)       | x: vector containing data of interest |
| Median                    | median(x)     | x: vector containing data of interest |
| sample standard deviation | sd(x)         | x: vector containing data of interest |
| sample variation          | var(x)        | x: vector containing data of interest |
| range                     | range(x)      | x: vector containing data of interest |
| Five-number summary       | summary(x)    | x: vector containing data of interest |

**Data Visualization**

| Statistics Concept   | Function in R | parameters                            |
|--------------------|------------------|----------------------------------|
| histogram            | hist(x)       | x: vector containing data of interest |
| box-and-whisker plot | boxplot(x)    | x: vector containing data of interest |
|                      |               | x: vector containing data of interest |

**Probability Distribution**

| Statistics Concept     | Function in R                                  | Output                                                             |
|------------------|-----------------------|-------------------------------|
| Normal distribution    | pnorm(x, mean = mean, sd = standard deviation) | probability of a value in the sample being less than or equal to x |
| Binomial Distriubution | dbinom()                                       | probability of a value being equal to x                            |
|                        | pbinom()                                       |                                                                    |
| Geometric Distribution | dgeom()                                        |                                                                    |
|                        | pgeom()                                        |                                                                    |

**Hypothesis Testing**

| Statistics Concept | Function in R                                             |
|-------------------------|----------------------------------------------|
| z-test             | z.test(data, mu = population mean, sigma = population sd) |
| t-test             | t.test(data)                                              |

# Other Resources

-   <https://www.w3schools.com/r/r_stat_intro.asp>
